Compare scANVI and SingleR annotations

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
update section where we look at overall agreement

# palette files
recoded_palette_file <- file.path(
  module_dir,
  "palettes",
  "harmonized-cell-type-palette.tsv"
)
nbatlas_palette_file <- file.path(
  module_dir,
  "palettes",
  "nbatlas-plus-cell-type-palette.tsv"
)

@@

```{r fig.height=12, fig.width=15}
scanvi_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
  dplyr::add_count(library_id) |>
  dplyr::mutate(
    library_id = glue::glue("{library_id} (N={n})"),
    library_id = factor(library_id), 
    library_id = forcats::fct_infreq(library_id), 
    library_id = forcats::fct_rev(library_id)
  ) |>
  ############ into ggplot ################
  ggplot() + 
  aes(x = scanvi_posterior) + 
  # ugly but _very visible_ color
  geom_histogram(bins = 100, fill = "magenta", color = "magenta") + 
  facet_wrap(vars(library_id), scales = "free_y", ncol = 7)
```

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there ARE repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode consensus NAs to "Unknown"
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation), "Unknown", validation_group_annotation
    )
  ) |>
  dplyr::select(-consensus_annotation) |>
  # merge in the annotations
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id") |>
  # make it tidy
  tidyr::pivot_longer(
    c(validation_group_annotation, singler, scanvi), 
    names_to = "celltype_method", 
    values_to = "label"
  ) |>
  harmonize_celltypes(label, label_recoded) |>
  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label))
```

@@

```{r, fig.height = 11, fig.width = 11} 
# make a wide version for the heatmap
celltype_wide_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  dplyr::select(-label_recoded) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label
  )
make_jaccard_heatmap(
  celltype_wide_df,
  "scanvi",
  "singler",
  "scANVI label",
  "SingleR label"
)
```

@@

```{r, fig.width = 10, fig.height = 4}
frac_agree_df <- celltype_wide_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode")) |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(
    frac_agree = sum(scanvi == singler)/dplyr::n(), 
    library_size = dplyr::n()
  )
frac_agree_df <- colData(merged_sce) |> 
  as.data.frame() |> 
  dplyr::select(library_id, is_xenograft) |> 
  unique() |>
  dplyr::right_join(frac_agree_df, by = "library_id")

p1 <- ggplot(frac_agree_df) + 
  aes(x = frac_agree) + 
  geom_histogram() + 
  xlab("Fraction of SingleR and scANVI cells\nthat agree, per library")

p2 <- ggplot(frac_agree_df) + 
  aes(x = library_size, y = frac_agree, color = is_xenograft) + 
  geom_point() +
  labs(
    x = "Library size (n cells)",
    y = "Fraction of SingleR and scANVI\ncells that agree"
  ) +
  theme(legend.position = "bottom")

p1 + p2

>>>>

# palette files
recoded_palette_file <- file.path(
  module_dir,
  "palettes",
  "harmonized-cell-type-palette.tsv"
)
nbatlas_palette_file <- file.path(
  module_dir,
  "palettes",
  "nbatlas-cell-type-palette.tsv"
)

@@

```{r fig.height=12, fig.width=15}
scanvi_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode")) |>
  dplyr::add_count(library_id) |>
  dplyr::mutate(
    library_id = glue::glue("{library_id} (N={n})"),
    library_id = factor(library_id), 
    library_id = forcats::fct_infreq(library_id), 
    library_id = forcats::fct_rev(library_id)
  ) |>
  ############ into ggplot ################
  ggplot() + 
  aes(x = scanvi_posterior) + 
  # ugly but _very visible_ color
  geom_histogram(bins = 100, fill = "magenta", color = "magenta") + 
  facet_wrap(vars(library_id), scales = "free_y", ncol = 7)
```

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there ARE repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode consensus NAs to "Unknown"
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation), "Unknown", validation_group_annotation
    )
  ) |>
  dplyr::select(-consensus_annotation) |>
  # merge in the annotations
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id") |>
  # make it tidy
  tidyr::pivot_longer(
    c(validation_group_annotation, singler, scanvi), 
    names_to = "celltype_method", 
    values_to = "label"
  ) |>
  harmonize_celltypes(label, label_recoded) |>
  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label)) |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)
```

@@

```{r, fig.height = 11, fig.width = 11} 
# make a wide version for the heatmap
celltype_wide_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  dplyr::select(-label_recoded) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label
  )

make_jaccard_heatmap(
  celltype_wide_df,
  "scanvi",
  "singler",
  "scANVI label",
  "SingleR label"
)
```

@@

> I think it would be helpful to get a table in here that shows you % of each library that has exact agreement between SingleR and scANVI, % of each library where cells are in the same family (e.g., the same cell is labeled as two different T cell types), and % of each library that disagrees and is not in the same family.

In this case, we consider these families specifically:

* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
* Conventional dendritic cells (cDC): `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
* Myeloid: `Neutrophil` and `Macrophage`

```{r}
# celltype_wide_df uses the original labels
identical_agree_df <- celltype_wide_df |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(
    frac_identical = sum(scanvi == singler)/dplyr::n(), 
    library_size = dplyr::n()
  )

# create a wide data frame with "family" labels where possible
celltype_family_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  # create label_family grouping
  dplyr::mutate(
    label_family = dplyr::case_when(
      label %in% nbatlas_t_celltypes ~ "T cell",
      label %in% nbatlas_nk_celltypes ~ "NK cell",
      label %in% nbatlas_cdc_celltypes ~ "cDC",
      label %in% nbatlas_monocyte_celltypes ~ "Monocyte",
      label %in% nbatlas_myeloid_celltypes ~ "Myeloid", 
      .default = label
    )
  ) |>
  dplyr::select(-label_recoded, -label) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label_family
  )

family_agree_df <- celltype_family_df |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(frac_same_family = sum(scanvi == singler)/dplyr::n())

# if families disagree, then there is no agreement at all
disagree_df <- celltype_family_df |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(frac_disagree = sum(scanvi != singler)/dplyr::n())
```

```{r}
# combine into a single table for display & viz
frac_df <- identical_agree_df |>
  dplyr::left_join(family_agree_df) |>
  dplyr::left_join(disagree_df) |>
  # round to 4 decimals
  dplyr::mutate(dplyr::across(contains("frac_"), round, 4)) |>
  dplyr::select(
    library_id, 
    library_size, 
    frac_identical, 
    frac_same_family, 
    frac_disagree
  ) |>
  dplyr::arrange(library_size)

# use DT for improved exploration ability
DT::datatable(frac_df, rownames = FALSE)
```

```{r}
# add PDX column for plotting
frac_df <- celltype_df |> 
  dplyr::select(library_id, is_xenograft) |>
  unique() |>
  dplyr::right_join(frac_df)

ggplot(frac_df) + 
  aes(x = library_size, y = frac_same_family) + 
  geom_point(aes(color = is_xenograft)) + 
  geom_smooth(method = "lm") +
  labs(
    x = "Library size",
    y = "Fraction of SingleR/scANVI labels from same family"
  ) +
  # stop going above 1
  scale_y_continuous(limits = c(0, 1))
```

Most cells have agreeing annotations in the majority of libraries.
There appears to be a slight trend (not signficant) that larger libraries have more agreement between methods, but this also seems driven by the libraries with low correspondence.

### Libraries with high levels of disagreement

There are four libraries (2 PDX and 2 tissue) which show less than 50% agreement, so we should look at these a bit more closely.
Notably, these do _not_ correspond to the four smallest libraries.
